Metacircular Virtualization \& \\ Practical Nock Interpretation

### 2025-07-02 mss.tex
Post

\begin{abstract}
\end{abstract}

@@

Given an argument for

Some essential components provided by vanilla Hoon for virtualizing Nock include:

@@

\noindent
\lstinline[style=inlinecode]{++edit} traverses the noun to attempt to access a given axis; if it succeeds, it returns a new noun with the value at that axis replaced by the given value.  If it fails, it returns an empty \texttt{unit}, indicating without a crash that the edit failed.

@@

Much of consequence is entailed by the use of a \lstinline[style=inlinecode]{scry} gate, which is a Hoon function that retrieves a value from an associated \lstinline[style=inlinecode]{sky} namespace (nominally, the bound scry namespace, although this is not necessarily entailed by Urbit in contemporary practice).  The \lstinline[style=inlinecode]{scry} function bears the signature \lstinline[style=inlinecode]{$-(^ (unit (unit)))}.  The result of this operation is either a noun or a block, and it is returned as a \lstinline[style=inlinecode]{(unit (unit noun))}.  These correspond to standard scry results in this order:

@@

\begin{lstlisting}[style=listingcode]
++  roof
  |$  [vase]
  $-  $:  lyc=(unit (set ship))     ::  leakset
          pov=path                  ::  provenance
          [vis=view bem=beam]       ::  perspective
      ==                            ::
  %-  unit                          ::  ~: unknown
  %-  unit                          ::  ~ ~: invalid
  (pair mark vase)                  ::  envased result
+$  view  $@(term [way=term car=term]) ::  perspective
\end{lstlisting}

@@

\lstinline[style=inlinecode]{++look} is a gate that is used to convert the \lstinline[style=inlinecode]{++mink}-compatible version of the scry handler into a version that can be used by the Arvo core.  This gate acts as a bridge between the \lstinline[style=inlinecode]{.^} dotket and the scry handler (\lstinline[style=inlinecode]{++look}) with access to the roof that came from the vane.  This pattern imposes constraints on interpreters, but provides a way to mediate access to the state of the system in a way that is consistent with Nock semantics.   Thus \lstinline[style=inlinecode]{.^} dotket gives mediated access into the \lstinline[style=inlinecode]{sky} without exposing the state of the entire system in the subject (untyped permissions-free access).

@@

\lstinline[style=inlinecode]{++mink} is naturally jetted for efficiency, ultimately by \lstinline[style=inlinecode]{u3m_soft_m()} in Urbit's Vere (C) and by \lstinline[style=inlinecode]{interpret()} in NockApp's Sword (Rust), both the virtualization context for their respective runtimes.  Some internal affordances for the runtime are in place to keep execution consistent; e.g., scry gates are stored on a stack to prevent re-entry if it scries.

TODO

Finally, it is instructive to consider what the minimal required amount of code machinery would be for one to successfully implement a Nock metacircular interpreter in Nock itself.  An examination of \lstinline[style=inlinecode]{++mink} shows modest reliance on Hoon's higher-level features; tree math, for instance, in \lstinline[style=inlinecode]{++cap} for distinguishing head vs.\ tail and \lstinline[style=inlinecode]{++mas} for finding the relative address of an axis.  (None of these are particularly complicated, but it seems much more straightforward to start from a compiled Hoon implementation and cut that down.)

TODO

@@

Metacircular virtualization is a powerful tool for building interpreters, and Nock is no exception.  The design of \lstinline[style=inlinecode]{++mock} and its underlying \lstinline[style=inlinecode]{++mink} function provides a practical way to interpret Nock in a way that is both efficient and extensible.  One could imagine, for instance, a Nock interpreter cousin to \lstinline[style=inlinecode]{++mock} that is instrumented for partial evaluation of formulas (as in a debugger or syntax highlighter).  Nock supports a rich capacity to build metacircular interpreters as a first-class feature, a potential which remains as yet relatively unexplored in the Nock ecosystem.

@@

>>>>

\begin{abstract}
  This paper presents a detailed examination of metacircular virtualization in the Nock combinator calculus, focusing on the practical implementation of a virtualized Nock interpreter within the Urbit ecosystem. We explore the design and functionality of key components, including the \lstinline[style=inlinecode]{++mink} interpreter, and their roles in enabling robust Nock execution environments.  Extensions of Nock using fake opcodes, particularly opcode 12, are discussed, and a minimal working example of a Nock-virtualized Nock formula is provided.
\end{abstract}

@@

Given an argument for metacircularity, we can now examine how Nock is actually virtualized in practice.  The Urbit standard library provides a metacircular Nock interpreter, \lstinline[style=inlinecode]{++mink}, along with several other components for building virtualized Nock environments.  Some essential components provided by vanilla Hoon for virtualizing Nock include:

@@

\noindent
\lstinline[style=inlinecode]{++edit} traverses the noun to attempt to access a given axis; if it succeeds, it returns a new noun with the original value at that axis replaced by a new target value.  If it fails, it returns an empty \texttt{unit}, indicating without a crash that the edit failed.

@@

\sloppy
Much of consequence is entailed by the use of a \lstinline[style=inlinecode]{scry} gate, which is a Hoon function that retrieves a value from an associated \lstinline[style=inlinecode]{sky} namespace (nominally, the bound scry namespace, although this is not necessarily entailed by Urbit in contemporary practice).  The \lstinline[style=inlinecode]{scry} function bears the signature
\begin{lstlisting}[style=listingblock]
$-(^ (unit (unit)))
\end{lstlisting}

\noindent
The result of this operation is either a noun or a block, and it is returned as a \lstinline[style=inlinecode]{(unit (unit noun))}.  These correspond to standard scry results in this order:

@@

\begin{lstlisting}[style=listingcode]
++  roof
  |$  [vase]
  $-  $:  lyc=(unit (set ship))     ::  leakset
          pov=path                  ::  provenance
          [vis=view bem=beam]       ::  perspective
      ==                            ::
  %-  unit                          ::  ~      unknown
  %-  unit                          ::  ~ ~    invalid
  (pair mark vase)                  ::  envased result
+$  view  $@(term [way=term car=term]) ::  perspective
\end{lstlisting}

@@

\lstinline[style=inlinecode]{++look} is a gate that is used to convert the \lstinline[style=inlinecode]{++mink}-compa\-tible version of the scry handler into a version that can be used by the Arvo core.  This gate acts as a bridge between the \lstinline[style=inlinecode]{.^} dotket and the scry handler (\lstinline[style=inlinecode]{++look}) with access to the roof that came from the vane.  This pattern imposes constraints on interpreters, but provides a way to mediate access to the state of the system in a way that is consistent with Nock semantics.   Thus \lstinline[style=inlinecode]{.^} dotket gives mediated access into the \lstinline[style=inlinecode]{sky} without exposing the state of the entire system in the subject (untyped permissions-free access).

@@

\begin{lstlisting}[style=listingcode]
  [^ *]
=/  head  $(formula -.formula)
?.  ?=(%0 -.head)  head
=/  tail  $(formula +.formula)
?.  ?=(%0 -.tail)  tail
[%0 product.head product.tail]
\end{lstlisting}

\noindent
There is little to be said of this machinery, but particularly notice that the error is propagated from the member of the cell which crashes.

\sloppy
\lstinline[style=inlinecode]{++mink} is naturally jetted for efficiency, ultimately by \lstinline[style=inlinecode]{u3m_soft_run()} in Urbit's Vere (C) and by \lstinline[style=inlinecode]{interpret()} in NockApp's NockVM (n\'{e}e) Sword (Rust), both the virtualization context for their respective runtimes.  Some internal affordances for the runtime are in place to keep execution consistent; e.g., scry gates are stored on a stack to prevent re-entry if it scries.

Nock is a crash-only language, which means that when something goes wrong, the exception is an exception for the runtime or virtual machine to handle as it will.  Typically, for a runtime this means an injection of the error trace back into the Nock kernel (cf.\ section ``\texttt{u3n}: nock execution'' in \texttt{u3.md} in the Urbit documentation\footnote{This section devoted a fair amount of attention to opcode 11 now 12 (it being written for Nock 5K) and hint evaluation.}).  (For \lstinline[style=inlinecode]{++mink}, errors are attempted to be front-run by the interpreter and thus bad Nock should not be executed as such.)

Hearkening back to \citet{Amin2017}'s collapsed towers of interpreters, the affordances of \lstinline[style=inlinecode]{++mock} as a metacircular interpreter permit us to build a tower of interpreters on top of it without incurring excessive performance overhead.  As the author of \texttt{u3.md} concluded, ``we [even] simply treat Nock proper as a special case of \lstinline[style=inlinecode]{mock}.''

What would such an interpreter look like, and how would it relate to actual operation on a runtime in practice?  Let us extend \lstinline[style=inlinecode]{++mink} to implement a logical loobean operator.  (Call the modified gate \lstinline[style=inlinecode]{++pink}.)  This operation will be invoked with a new fake opcode 13.  Under the hood, opcode 13 will simply supply logically negate a value (via an opcode 6), much as how opcode 12 supplies a consistent Nock formula for the scry operation.  (Similar to \lstinline[style=inlinecode]{[0 0]}, we will be able to use \lstinline[style=inlinecode]{[13 1 2]} to initiate a crash if desired.)

\begin{lstlisting}[style=listingcode]
:: *[a 13 b]      NOT *[a b]
  [%13 subject=*]
=/  argument  $(formula argument.formula)
?.  ?=(%0 -.argument)  argument
?^  product.argument  [%2 trace]
?:  =(%1 product.argument)  [%0 %.y]
?:  =(%0 product.argument)  [%0 %.n]
[%2 trace]
\end{lstlisting}

This is invoked as:

\begin{lstlisting}[style=listingcode]
> (pink [0 13 1 0] *$-(^ (unit (unit))))
[%0 product=0]
> (pink [0 13 1 1] *$-(^ (unit (unit))))
[%0 product=1]
> (pink [0 13 1 2] *$-(^ (unit (unit))))
[%2 trace=~]
\end{lstlisting}

Finally, it is instructive to consider what the minimal required amount of code machinery would be for one to successfully implement a Nock metacircular interpreter in Nock itself.  An examination of \lstinline[style=inlinecode]{++mink} shows modest reliance on Hoon's higher-level features; tree math, for instance, in \lstinline[style=inlinecode]{++cap} for distinguishing head vs.\ tail and \lstinline[style=inlinecode]{++mas} for finding the relative address of an axis.  (Neither of these are particularly complicated, but they entail some basic arithmetic operations as well.)  By omitting \lstinline[style=inlinecode]{++scry} and opcode 12, as well as reducing to a structural match with a \lstinline[style=inlinecode]{+$tone} instead of an explicit Hoon type match, one could produce a reasonably sized Hoon program compiling to something close to the minimal viable Nock virtualized interpreter.  No doubt intrepid code golfers will find it very much an upper bound rather than a lower.

This Nock formula produces a working virtualized Nock interpreter derived from Hoon code with some hand-tuned design.  The input function signature is \lstinline[style=inlinecode]{[subject=* formula=*]} with output of the form \lstinline[style=inlinecode]{[@ product=*]}; if the head is \lstinline[style=inlinecode]{%0}, then the product is a valid Nock noun, while a head of \lstinline[style=inlinecode]{%bail} indicates a crash in the evaluation.  Opcode 9 calls may be addressed into the core as subject, and the formula itself is complete with an empty subject.

\begin{lstlisting}[style=listingcode]
[1
 :: +mul multiplication of two atoms
 [8 [1 1 1]
    [1 8 [1 0]
       8 [1 6 [5 [1 0] 0 60]
              [0 6]
              9 2 10 [60 8 [9 686 0 31]
                            9 2 10 [6 0 124] 0 2]
                  10 [6 8 [9 20 0 31]
                            9 2 10 [6 [0 125] 0 14]
                      0 2] 0 1]
     9 2 0 1] 0 1]
 [:: +add addition of two atoms
  [8 [1 0 0]
     [1 6 [5 [1 0] 0 12]
          [0 13]
          9 2 10
          [6 [8 [9 686 0 7] 9 2 10 [6 0 28] 0 2]
           4 0 13]
        0 1] 0 1]
    [:: +list type definition
      [8
        [8 [1 0]
           [1 8 [0 6] 8 [5 [0 14] 0 2] 0 6]
         0 1]
        [1 8 [1 0]
          [1 8
             [6 [3 0 6]
                [[6 [5 [1 0] 0 12]
                    [1 0]
                    0 0]
                 8 [0 30] 9 2 10 [6 0 29] 0 2]
              6 [5 [1 0] 0 6]
                [1 0]
                0 0]
            8 [5 [0 14] 0 2] 0 6]
          0 1] 0 1]
      [:: +dvr division with remainder
       8 [1 1 1]
         [1 6 [5 [1 0] 0 13]
              [0 0]
              8 [1 0]
              8 [1 6 [8 [9 687 0 31]
                      9 2 10 [6 [0 124] 0 125] 0 2]
                 [[0 6] 0 60]
                 9 2 10
                 [60 8 [9 23 0 31]
                       9 2 10 [6 [0 124] 0 125] 0 2]
                 10 [6 4 0 6] 0 1]
           9 2 0 1] 0 1]
      [:: +cap: 
       8 [1 0]
         [1 6 [5 [1 2] 0 6]
              [1 2]
              6 [5 [1 3] 0 6]
                [1 3]
                6 [6 [5 [1 1] 0 6] [1 0] 5 [1 0] 0 6]
                  [0 0]
                  9 2 10
                  [6 7 [8 [9 170 0 7]
                           9 2 10
                           [6 [0 14] 7 [0 3] 1 2]
                        0 2] 0 2]
          0 1] 0 1]
      [:: +dec decrement an atom
       8 [1 0]
         [1 6 [5 [1 0] 0 6]
              [0 0]
              8 [1 0]
              8 [1 6 [5 [0 30] 4 0 6]
                     [0 6]
                     9 2 10 [6 4 0 6] 0 1]
          9 2 0 1] 0 1]
      :: +lth less-than test
      8 [1 0 0]
        [1 6 [6 [5 [0 12] 0 13] [1 1] 1 0]
             [6 [8
                 [1 6 [5 [1 0] 0 28]
                      [1 0]
                      6 [6 [6 [5 [1 0] 0 29]
                              [1 1]
                              1 0]
                           [6 [9 2 10
                               [14 [8 [9 686 0 15]
                                    9 2 10 [6 0 60]
                                    0 2]
                                8 [9 686 0 15]
                                9 2 10 [6 0 61] 0 2]
                               0 1]
                            [1 0]
                            1 1]
                          1 1]
                   [1 0] 1 1] 9 2 0 1]
          [1 0] 1 1] 1 1] 0 1]
    [:: +jink, raw virtual Nock
     8 [1 0 0]
       [1 8
          [1 [:: +frag fragmentary subtree of noun
              8 [1 0 0]
                [1 6 [5 [1 0] 0 12]
                     [1 0]
                   8 [1 6 [5 [1 1] 0 28]
                          [[1 0] 0 29]
                          6 [6 [3 0 29] [1 1] 1 0]
                            [1 0]
              8 [8 [9 342 0 127] 9 2 10 [6 0 60] 0 2]
                9 2 10
                [14 [8 [9 87 0 255]
                       9 2 10 [6 0 124] 0 2]
                  6 [5 [1 2] 0 2]
                    [0 122]
                    6 [5 [1 3] 0 2]
                      [0 123]
                      1 1.818.845.538 0]
                 0 3] 9 2 0 1] 0 1]
             [:: +edit edit an atom at a given axis
              8 [1 0 0 0]
                [1 6 [5 [1 1] 0 12]
                     [[1 0] 0 27]
                     6 [6 [3 0 26] [1 1] 1 0]
                     [1 0]
              8 [9 2 10
                 [26
                  8 [8 [9 342 0 63]
                     9 2 10 [6 0 28] 0 2]
                    6 [5 [1 2] 0 2]
                      [0 116]
                      6 [5 [1 3] 0 2]
                        [0 117]
                        0 0]
                 10 [12 8 [9 87 0 127]
                    9 2 10 [6 0 60] 0 2] 0 3]
              6 [5 [1 0] 0 2]
                [1 0]
                6 [5 [1 2] 0 6]
                  [[1 0] [0 5] 0 245]
                  6 [5 [1 3] 0 6]
                    [[1 0] 0 244]
                    1 1.818.845.538 0]
              0 5] 0 1]
          6  :: Cell distribution
          [6 [3 0 29] [3 0 58] 1 1]
          [8 [9 11 10 [29 0 58] 0 1]
            6 [5 [1 0] 0 4]
              [8 [9 11 10 [29 0 123] 0 3]
                6 [5 [1 0] 0 4]
                  [[1 0] [0 13] 0 5]
                  0 2]
              0 2]
          6  :: Opcode 0
          [6 [3 0 29]
             [6 [5 [1 0] 0 58]
                [6 [3 0 59] [1 1] 1 0]
                1 1]
             1 1]
          [8 [8 [9 4 0 1]
                9 2 10 [6 [0 123] 0 60] 0 2]
             6 [5 [1 0] 0 2]
               [1 1.818.845.538 0]
               [1 0]
             0 5]
          6  :: Opcode 1
          [6 [3 0 29]
             [5 [1 1] 0 58] 1 1]
          [[1 0] 0 59]
          6  :: Opcode 2
          [6 [3 0 29]
             [6 [5 [1 2] 0 58] [3 0 59] 1 1] 1 1]
          [8 [9 11 10 [29 0 118] 0 1]
             6 [5 [1 0] 0 4]
             [8
              [9 11 10 [29 0 247] 0 3]
              6
              [5 [1 0] 0 4]
              [9 11 10 [14 [0 13] 0 5] 0 7]
              0 2]
            0 2]
          6  :: Opcode 3
          [6 [3 0 29]
             [5 [1 3] 0 58] 1 1]
          [8 [9 11 10 [29 0 59] 0 1]
           6 [5 [1 0] 0 4]
             [[1 0] 3 0 5]
             0 2]
          6  :: Opcode 4
          [6 [3 0 29]
             [5 [1 4] 0 58] 1 1]
          [8
            [9 11 10 [29 0 59] 0 1]
            6
            [5 [1 0] 0 4]
            [6 [6 [3 0 5] [1 1] 1 0]
               [[1 0] 4 0 5]
               1 1.818.845.538 0]
            0 2]
          6  :: Opcode 5
          [6 [3 0 29]
             [6 [5 [1 5] 0 58] [3 0 59] 1 1] 1 1]
          [ 8
            [9 11 10 [29 0 118] 0 1]
            6
            [5 [1 0] 0 4]
            [ 8
              [9 11 10 [29 0 247] 0 3]
              6
              [5 [1 0] 0 4]
              [[1 0] 5 [0 13] 0 5]
              0 2]
            0 2]
          6  :: Opcode 6
          [6 [3 0 29]
             [6 [5 [1 6] 0 58]
                [6 [3 0 59] [3 0 119] 1 1]
                1 1] 1 1]
          [8 [9 11 10 [29 0 118] 0 1]
             6
             [5 [1 0] 0 4]
             [6 [5 [1 0] 0 5]
                [9 11 10 [29 0 494] 0 3]
                6 [5 [1 1] 0 5]
                  [9 11 10 [29 0 495] 0 3]
                  1 1.818.845.538 0]
::              11
::              [1.936.945.012 1 0]
::              1 1.818.845.538 0
            0 2]
          6  :: Opcode 7
          [6 [3 0 29]
             [6 [5 [1 7] 0 58] [3 0 59] 1 1] 1 1]
          [8 [9 11 10 [29 0 118] 0 1]
             6 [5 [1 0] 0 4]
               [9 11 10 [14 [0 5] 0 247] 0 3]
               0 2]
          6  :: Opcode 8
          [6 [3 0 29]
             [6 [5 [1 8] 0 58] [3 0 59] 1 1] 1 1]
          [8 [9 11 10 [29 0 118] 0 1]
             6 [5 [1 0] 0 4]
               [9 11 10 [14 [[0 5] 0 60] 0 247] 0 3]
               0 2]
          6  :: Opcode 9
          [6 [3 0 29]
             [6 [5 [1 9] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118] [1 1] 1 0]
                   1 1]
                1 1]
             1 1]
          [8 [9 11 10 [29 0 119] 0 1]
             6 [5 [1 0] 0 4]
               [8 [8 [9 4 0 3]
                     9 2 10 [6 [0 502] 0 13] 0 2]
               6 [5 [1 0] 0 2]
                 [1 1.818.845.538 0]
              9 11 10 [14 [0 13] 0 5]
              0 7]
            0 2]
          6  :: Opcode 10
          [6 [3 0 29]
             [6 [5 [1 10] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118]
                      [6 [3 0 236] [1 1] 1 0]
                      1 1]
                   1 1]
                1 1]
             1 1]
          [6 [5 [1 0] 0 236]
             [1 1.818.845.538 0]
             8 [9 11 10 [29 0 119] 0 1]
             6 [5 [1 0] 0 4]
               [8 [9 11 10 [29 0 493] 0 3]
               6 [5 [1 0] 0 4]
               [8 [8 [9 10 0 7]
                     9 2 10 [6 [0 2.028] [0 29] 0 13]
                   0 2]
                6 [5 [1 0] 0 2]
                  [1 1.818.845.538 0]
                  [1 0]
                0 5]
               0 2]
             0 2]
          6  :: Opcode 11
          [6 [3 0 29]
             [6 [5 [1 11] 0 58]
                [6 [3 0 59]
                   [6 [3 0 118]
                      [1 1]
                      1 0]
                   1 1]
                1 1]
             1 1]
          [8 [9 11 10 [29 0 119] 0 1]
             6 [5 [1 0] 0 4]
               [[1 0]
                2 [0 60] [1 11] [0 246] [1 1] 0 5]
            0 2]
::          11
::          [1.936.945.012 1 0]
          1 1.818.845.538 0]
        9 11 0 1]
      0 1]
    :: +mas
    8 [1 0]
      [1 6 [6 [5 [1 3] 0 6] [1 0] 5 [1 2] 0 6]
           [1 1]
           6 [6 [5 [1 1] 0 6] [1 0] 5 [1 0] 0 6]
             [0 0]
             8 [9 20 0 7]
             9 2 10
             [6 [7 [0 3] 7 [8 [9 170 0 7]
                            9 2 10
                            [6 [0 14] 7 [0 3] 1 2]
                            0 2] 0 3]
        7 [0 3] 8 [9 4 0 7]
        9 2 10 [6
          [7 [0 3]
            9 2 10
            [6 7 [8 [9 170 0 7]
                  9 2 10 [6 [0 14] 7 [0 3] 1 2]
                  0 2] 0 2]
            0 1]
          7 [0 3] 1 2]
        0 2]
      0 2]
    0 1]
  [:: +unit, type definition of unit/Maybe
   8
    [8 [1 0] [1 8 [0 6] 8 [5 [0 14] 0 2] 0 6] 0 1]
    [1 8 [1 0]
         [1 8 [6 [3 0 6]
                 [[8 [0 30] 9 2 10 [6 0 28] 0 2]
                  8 [7 [0 7] 8 [9 22 0 7]
                       9 2 10 [6 0 14] 0 2]
                  9 2 10 [6 0 29] 0 2]
               6 [5 [1 0] 0 6]
                 [1 0]
                 0 0]
          8 [5 [0 14] 0 2]
          0 6]
      0 1]
    0 1]
  :: +sub subtraction of two atoms
  8 [1 0 0]
    [1 6 [5 [1 0] 0 13]
         [0 12]
         9 2 10
         [6 [8 [9 686 0 7] 9 2 10 [6 0 28] 0 2]
          8 [9 686 0 7] 9 2 10 [6 0 29] 0 2]
    0 1] 0 1]
\end{lstlisting}

TODO there's a missing open bracket I need to chase down

@@

Metacircular virtualization is a powerful tool for building interpreters, and Nock is not only no exception, it exemplifies the practice.  The design of \lstinline[style=inlinecode]{++mock} and its underlying \lstinline[style=inlinecode]{++mink} function provides a practical way to interpret Nock in a way that is both efficient and extensible.  One could imagine, for instance, a Nock interpreter cousin to \lstinline[style=inlinecode]{++mock} that is instrumented for partial evaluation of formulas (as in a debugger or syntax highlighter robust to syntax errors).  Nock supports a rich capacity to build metacircular interpreters as a first-class feature, a potential which remains as yet relatively unexplored in the Nock ecosystem.

@@

% ::
% ::  +jink, a search for the smallest possible +mock Nock virtualized interpreter
% ::
% =<
%   :-  (jink [42 [0 1]])
%   (jink [42 [0 0]])
% |%
% ::
% ++  unit
%   |$  [item]
%   $@(~ [~ u=item])
% ::
% ++  list
%   |$  [item]
%   $@(~ [i=item t=(list item)])
% ::
% ++  cap
%   |=  a=@
%   ^-  ?(%2 %3)
%   ?-  a
%     %2        %2
%     %3        %3
%     ?(%0 %1)  !!
%     *         $(a -:(dvr a 2))
%   ==
% ::
% ++  mas
%   |=  a=@
%   ^-  @
%   ?-  a
%     ?(%2 %3)  1
%     ?(%0 %1)  !!
%     *         (add +:(dvr a 2) (mul $(a -:(dvr a 2)) 2))
%   ==
% ::
% ++  add
%   |=  [a=@ b=@]
%   ^-  @
%   ?:  =(0 a)  b
%   $(a (dec a), b +(b))
% ::
% ++  dec
%   |=  a=@
%   ?<  =(0 a)
%   =+  b=0
%   |-  ^-  @
%   ?:  =(a +(b))  b
%   $(b +(b))
% ::
% ++  dvr
%   |:  [a=`@`1 b=`@`1]
%   ^-  [p=@ q=@]
%   ?<  =(0 b)
%   =+  c=0
%   |-
%   ?:  (lth a b)  [c a]
%   $(a (sub a b), c +(c))
% ::
% ++  lth
%   |=  [a=@ b=@]
%   ^-  ?
%   ?&  !=(a b)
%       |-
%       ?|  =(0 a)
%           ?&  !=(0 b)
%               $(a (dec a), b (dec b))
%   ==  ==  ==
% ::
% ++  mul
%   |:  [a=`@`1 b=`@`1]
%   ^-  @
%   =+  c=0
%   |-
%   ?:  =(0 a)  c
%   $(a (dec a), c (add b c))
% ::
% ++  sub
%   |=  [a=@ b=@]
%   ^-  @
%   ?:  =(0 b)  a
%   $(a (dec a), b (dec b))
% ::
% ::  +jink: raw virtual nock
% ::
% ++  jink
%   |=  [subject=* formula=*]
%   |^  ^-  [@ product=*]
%   ?+  formula  [%bail ~]
%       [^ *]
%     =/  head  $(formula -.formula)
%     ?.  ?=(%0 -.head)  head
%     =/  tail  $(formula +.formula)
%     ?.  ?=(%0 -.tail)  tail
%     [%0 product.head product.tail]
%     ::
%       [%0 axis=@]
%     =/  part  (frag axis.formula subject)
%     ?~  part  [%bail ~]
%     [%0 u.part]
%     ::
%       [%1 constant=*]
%     [%0 constant.formula]
%     ::
%       [%2 subject=* formula=*]
%     =/  subject  $(formula subject.formula)
%     ?.  ?=(%0 -.subject)  subject
%     =/  formula  $(formula formula.formula)
%     ?.  ?=(%0 -.formula)  formula
%     %=  $
%       subject  product.subject
%       formula  product.formula
%     ==
%     ::
%       [%3 argument=*]
%     =/  argument  $(formula argument.formula)
%     ?.  ?=(%0 -.argument)  argument
%     [%0 .?(product.argument)]
%     ::
%       [%4 argument=*]
%     =/  argument  $(formula argument.formula)
%     ?.  ?=(%0 -.argument)  argument
%     ?^  product.argument  [%bail ~]
%     [%0 .+(product.argument)]
%     ::
%       [%5 a=* b=*]
%     =/  a  $(formula a.formula)
%     ?.  ?=(%0 -.a)  a
%     =/  b  $(formula b.formula)
%     ?.  ?=(%0 -.b)  b
%     [%0 =(product.a product.b)]
%     ::
%       [%6 test=* yes=* no=*]
%     =/  result  $(formula test.formula)
%     ?.  ?=(%0 -.result)  result
%     ?+  product.result  [%bail ~]
%       %&  $(formula yes.formula)
%       %|  $(formula no.formula)
%     ==
%     ::
%       [%7 subject=* next=*]
%     =/  subject  $(formula subject.formula)
%     ?.  ?=(%0 -.subject)  subject
%     %=  $
%       subject  product.subject
%       formula  next.formula
%     ==
%     ::
%       [%8 head=* next=*]
%     =/  head  $(formula head.formula)
%     ?.  ?=(%0 -.head)  head
%     %=  $
%       subject  [product.head subject]
%       formula  next.formula
%     ==
%     ::
%       [%9 axis=@ core=*]
%     =/  core  $(formula core.formula)
%     ?.  ?=(%0 -.core)  core
%     =/  arm  (frag axis.formula product.core)
%     ?~  arm  [%bail ~]
%     %=  $
%       subject  product.core
%       formula  u.arm
%     ==
%     ::
%       [%10 [axis=@ value=*] target=*]
%     ?:  =(0 axis.formula)  [%bail ~]
%     =/  target  $(formula target.formula)
%     ?.  ?=(%0 -.target)  target
%     =/  value  $(formula value.formula)
%     ?.  ?=(%0 -.value)  value
%     =/  mutant=(unit *)
%       (edit axis.formula product.target product.value)
%     ?~  mutant  [%bail ~]
%     [%0 u.mutant]
%     ::
%       [%11 tag=@ next=*]
%     =/  next  $(formula next.formula)
%     ?.  ?=(%0 -.next)  next
%     :-  %0
%     .*  subject
%     [11 tag.formula 1 product.next]
%   ==
%   ::
%   ++  frag
%     |=  [axis=@ noun=*]
%     ^-  (unit)
%     ?:  =(0 axis)  ~
%     |-  ^-  (unit)
%     ?:  =(1 axis)  `noun
%     ?@  noun  ~
%     %=  $
%       axis  (mas axis)
%       noun  ?-((cap axis) %2 -.noun, %3 +.noun)
%     ==
%   ::
%   ++  edit
%     |=  [axis=@ target=* value=*]
%     ^-  (unit)
%     ?:  =(1 axis)  `value
%     ?@  target  ~
%     =/  mutant
%       %=  $
%         axis    (mas axis)
%         target  ?-((cap axis) %2 -.target, %3 +.target)
%       ==
%     ?~  mutant  ~
%     ?-  (cap axis)
%       %2  `[u.mutant +.target]
%       %3  `[-.target u.mutant]
%     ==
%   --
% --
